A Fast and Tight Heuristic for A* in Road Networks

### 2020-08-21 paper/ch_potentials.tex
shrink boxplot
stuff

\def\year{2021}\relax
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai20}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{graphicx}  % DO NOT CHANGE THIS
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS

@@

\pdfinfo{
/Title (A Fast and Tight Heuristic for A* in Road Networks)
/Author (Ben Strasser, Tim Zeitz)
/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
}
\title{A Fast and Tight Heuristic for A* in Road Networks}
\author{Ben Strasser\\
academia@ben-strasser.net\\
\And
Tim Zeitz\\
tim.zeitz@kit.edu\\
Karlsruhe Institute of Technology
}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
We call our technique CH-Potentials.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is often formalized as a variant of the classical shortest path problem in weighted graphs.
Developing flexible techniques to efficiently compute exact routes is a tough challenge.
\emph{Exact} means that the computed path always has minimum weight.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations or handle more complex weights (e.g. travel times depending on the time of the day) is \emph{flexibility}.

@@

\caption{Nodes explored by A* with CH-Potentials. Nodes were explored from blue through green to red.}
\label{img:search-space}
\end{figure}

Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
It iteratively explores nodes of the graph in ascending distance from the source node until the target is reached.
A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
It changes the order in which nodes are explored by using the distance from the source plus an estimate of the remaining distance to the target as the ordering criterion.
Figure~\ref{img:search-space} depicts an example A* search.
While flexible, A*'s performance crucially depends on the quality of the employed estimating heuristic.
Finding good heuristics that can be evaluated efficiently is difficult.

In this paper, we consider a setup with a preprocessing phase and a query phase.
The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with lower bound weights $w_\ell$.
The source $s$ and target $t$ of and the actual edge weights $w_q$ are the input to the query phase in which shortest paths are computed.
We require that $w_q(e)\ge w_\ell(e)$ for all edges $e \in E$.
%
Street networks of continental scale have many millions of nodes.
% While the preprocessing phase may be slow, it must eventually finish.
On graphs of this size, quadratic preprocessing time is prohibitive.

In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
The preprocessing phase computes auxiliary data to efficiently evaluate the heuristic.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
We call our algorithm \emph{CH-Potentials}, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.

Formally, a heuristic $h$ is a function that maps nodes onto numbers.
$h$ depends on the target $t$ of the search but not one the source $s$.
Without loss of generality, we assume that $h(t)=0$.
The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
CH-Potentials compute the tight heuristic with respect to $w_\ell$.
Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
In this sense, CH-Potentials are as tight as possible.

@@

There is a lot of work that extends CH to handle more complex settings.
For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In~\cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developed in~\cite{bdgwz-sfpcs-19}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
Our primary algorithm design objective is to retain A*'s flexibility.

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the tight heuristic.
Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
This is the tightest heuristic achievable during preprocessing.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

Many applications need flexibility.
The most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{dgpw-crprn-13,dsw-cch-15} and predicted traffic~\cite{ndls-bastd-12,bgsv-mtdtt-13}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}.
For example, for some users shorter routes are more important than fast routes.
Others want to avoid tunnels or highways.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
%
Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.

\section{Contribution \& Outline}

We introduce CH-Potentials, a two-phase technique to efficiently compute tight heuristics, in Section~\ref{sec:main-algo}.
Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we describe some improvements to A* to more efficiently process low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and how they can be solved using CH-Potentials.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
By comparing against Oracle-A*, we show that no alternative approach can be significantly faster, assuming that $w_\ell$ is the tightest lower bound that can be known in advance.

@@

A CH is a two phase technique to efficiently compute exact, shortest paths.
It is not flexible, i.e., $w_q=w_\ell$.
For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $v$ is more important than both its neighbors.
$v$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G+$ (blue) exists.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.

Using a CH, we can compute a tight A* heuristic.
In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
The query phase consists of an A* search, where the evaluation of the heuristic $\pi(v)$ performs a CH-query from $v$ to $t$.
%
This works, however, while one CH query is fast, answering one for every node explored in the $A^*$ search is slow.
Fortunately, we can do better.
However, for this we need another component called PHAST~\cite{dgnw-phast-13}.

\subsection{PHAST based heuristic}

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogue to the CH query:
From the target node $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the distances from all nodes towards $t$ is computed.
PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
PHAST's main advantage is that can easily be parallelized.
However, we will not consider parallelization in this paper.
We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can compute a tight A* heuristic in another way.
In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the lower bound edge weights $w_\ell$ and store the result in an array $H$.
Then, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

% This PHAST-based algorithm works.
The PHAST-based heuristic evaluation and by extension the $A^*$ search is indeed fast.
However, the PHAST step before the search is comparatively expensive.
The reason is that the distances towards $t$ are computed for \emph{all} nodes.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.

@@

Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
In a first step, we run the backward CH search from $t$ to obtain an array $B$.
$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

Suppose, we need to compute the potential $\pi(v)$ of a node $v$.
We do this by first recursively computing the potentials $\pi(u)$ of all nodes for which an up-edge $(v,u)$ exists.
Afterwards, we compute $d = \min_u\{w_\ell(v,u) + \pi(u)\}$ the minimum distance over all up-down paths that contain at least one up-edge.
As the shortest up-down path does not necessarily have to contain an up-edge, we set $\pi(v) = \min \{ B[v], d \}$.
This calculation is correct, as it computes the minimum up-down $vt$-path distance in $G^+$, which corresponds to the minimum $vt$-path distance in a CH.

The resulting algorithm is the basic CH-Potentials algorithm.
It is flexible and exact.
However, we can still improve the performance by reducing the number of queue operations and heuristic evaluations.

@@

Heuristic evaluation is an expensive operation that we want to avoid.
Together with the queue operations, it makes up a significant amount of the running time of A*.
The heuristic needs to be evaluated whenever a node is pushed into the queue.
As our experiments show, the total running time roughly corresponds to the number of queue pushs.
In this section, we propose optimizations to reduce the number of nodes pushed into the queue.

@@

Our algorithm differs from A* when removing a node $x$ from the queue.
A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
If A* succeeds, $y$'s weight in the queue is set to $D[y]+\pi(y)$.
Our algorithm, however, behaves differently, if $d(y)\le 2$.
Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
Instead, it iteratively tries to reduce all $D[y_i]$.
If it does not reach $z$, then only $D$ is modified but no queue action is performed.
If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+\pi(z)$.

@@

Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.

@@

A lot of nodes in road networks lead to dead-ends.
Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
This reduces the number of explored nodes and thus leads to fewer heuristic evaluations.

@@

A* is a very flexible algorithm.
Many extended problems can therefore be solved using the CH-Potentials algorithm.
In this section, we describe some extended problems and how they can be solved with CH-Potentials.

@@

Avoiding tunnels and/or highways is a common feature of navigation devices.
Implementing this feature with CH-Potentials is easy.
We set $w_\ell$ to the freeflow travel time.
If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
Otherwise, $w_q$ is set to the freeflow travel time.
This is all that is necessary to make our CH-Potentials algorithm avoid tunnels and/or highways.

@@

Truck routing differs from car routing due to night driving bans and other restrictions.
In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is successfully employed to efficiently find routes for trucks when edges may be blocked during certain times, drivers may need to wait at parking locations, and several routes with different trade-offs between arrival time and route quality may need to be considered.

>>>>

\def\year{2021}\relax
%File: formatting-instructions-latex-2021.tex
%release 2021.1
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai21}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS

@@

\pdfinfo{
/Title (A Fast and Tight Heuristic for A* in Road Networks)
/Author (Foo Bar)
/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
}
\title{A Fast and Tight Heuristic for A* in Road Networks}
\author{Foo Bar}

% \author{Ben Strasser\\
% academia@ben-strasser.net\\
% \And
% Tim Zeitz\\
% tim.zeitz@kit.edu\\
% Karlsruhe Institute of Technology
% }

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Efficiently computing flexible and exact routes in large road networks is a tough challenge.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
However, A*'s performance depends on the availability of a good heuristic.
Computing a good heuristic is a challenge of its own.
Shortest paths can also be quickly computed using hierarchical speedup techniques.
They are not based on A* and do not use a heuristic.
Hierarchical techniques are known to work well on road networks.
They achieve speed and exactness but sacrifice A*'s flexibility.
In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
We call our technique CH-Potentials.
With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
\end{abstract}

\section{Introduction}
\label{sec:intro}
Route planning in large street networks is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
It is often formalized as a variant of the classical shortest path problem in weighted graphs.
Developing flexible techniques to efficiently compute exact routes is a tough challenge.
\emph{Exact} means that a path of provable minimum weight is computed.
\emph{Efficiently} means that the path computation is fast.
Being able to change the edge weights between path computations is \emph{flexibility}.
Handling complex weights, such as edge weight functions, is another aspect of flexibility.

@@

\caption{Nodes explored by A* with CH-Potentials. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
\label{img:search-space}
\end{figure}

Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
It iteratively explores nodes ordered by their distance from the start.
A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
It processes nodes in a different order, preferring nodes closer to the destination.
Which nodes are considered close is encoded using a heuristic function.
While flexible, A*'s performance crucially depends on the quality of this function.
Finding good heuristics that can be evaluated efficiently is difficult.
Constructing good heuristics on road networks, such as the one depicted in Figure~\ref{img:search-space}, is the focus of this paper.

In this paper, we consider a setup with an offline preprocessing phase and an online query phase.
The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with \emph{lower bound weights} $w_\ell$.
The query phase is given the \emph{source} $s$ and the \emph{target} $t$ as input.
$s$ is also called start and synonyms for $t$ are goal or destination.
It is further given the \emph{query edge weights} $w_q$.
The task is to compute a minimum length path with respect to $w_q$.
Of $w_q$, it is only known that $w_\ell$ is a lower bound, i.e., $w_\ell \le w_q$.
%
While the preprocessing phase may be slow, it must eventually finish.
On graphs with millions of nodes, quadratic preprocessing time is prohibitive.

In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
In the preprocessing phase, it computes a Contraction Hierarchy (CH)~\cite{gssv-erlrn-12}.
The CH is used to efficiently evaluate the heuristic.
Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
We call our algorithm \emph{CH-Potentials}, following the ALT terminology.

A heuristic $h$ is a function that maps nodes onto numbers.
$h$ depends on the target $t$ but not on the source $s$.
Without loss of generality, we assume that $h(t)=0$.
The \emph{tight} heuristic assigns to every node $x$, the length of a shortest $xt$-path.
All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
CH-Potentials compute the tight heuristic with respect to $w_\ell$.
Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
In this sense, CH-Potentials are as tight as possible and thus optimal.

@@

There is a lot of work that extends CH to handle more complex settings.
For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
In~\cite{dsw-cch-15}, CH is modified to support Live-Traffic.
Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, a lot of past work uses A* to accelerate hierarchical approaches even further and loose A*'s' flexibility.

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
However, ALT's heuristic is not tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as tight heuristic.
Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
This results in a good but not a tight heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

To compare against related work, we introduce the concept of Oracle-A*.
It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
The array computation time is unaccounted for, i.e., it is magically filled instantly.
Oracle-A* is not achievable in practice.
However, it is a lower bound for every algorithm following our two phase setup.
We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

@@

Many applications need flexibility.
The most common ones are traffic avoidance, user preferences, and vehicle restrictions.
Traffic avoidance comes in two forms: Live~\cite{dgpw-crprn-13,dsw-cch-15} and predicted traffic~\cite{ndls-bastd-12,bgsv-mtdtt-13}.
Predicted traffic is often called time-dependent routing.
%
User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}.
For example, for some users shorter routes are more important than fast routes.
Others want to avoid tunnels or highways.
%
Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
%
\section{Contribution \& Outline}

We introduce CH-Potentials, a two-phase technique to efficiently compute tight heuristics, in Section~\ref{sec:main-algo}.
Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we improve A*'s handling of low degree nodes, common in road networks.
In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing algorithms for various extended route planning applications.
Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
By comparing against Oracle-A*, we show that no two-phase alternative exact A* approach can be significantly faster.

@@

A CH is a two phase technique to efficiently compute exact, shortest paths.
It is not flexible, i.e., requires $w_q=w_\ell$.
For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
In this section, we give an introduction.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G+$ (blue) exists.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

Using a CH, we can compute a simple and tight heuristic.
In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
The heuristic evaluation $h(x)$ performs a CH-query from $x$ to $t$.
%
This works, however, while one CH query is fast, answering one for every node explored in the $A^*$ search is slow.
Fortunately, we can do better.
However, for this we need another component called PHAST~\cite{dgnw-phast-13}.

\subsection{PHAST based Heuristic}

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
First the preprocessing phase is executed analogously to the original CH.
The query phase is divided into two steps.
The first step is analogue to the CH query:
From $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the distances from all nodes towards $t$ is computed.
%PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
%PHAST's main advantage is that can easily be parallelized.
%However, we will not consider parallelization in this paper.
%We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can also compute a tight A* heuristic.
In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

% This PHAST-based algorithm works.
The $H$ lookup and by extension the $A^*$ search is indeed fast.
However, the PHAST step before the search is comparatively expensive.
The reason is that the distances towards $t$ are computed for \emph{all} nodes.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.

@@

Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
In a first step, we run the backward CH search from $t$ to obtain an array $B$.
$B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

To compute the heuristic $h(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h(y)$.
Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h(y)\}$.
As not all shortest up-down paths contain an up-edge, we set $h(x) = \min \{ B[x], d \}$.
This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+$, which corresponds to the minimum $xt$-path distance in a CH.
A* with this heuristic is the basic CH-Potentials algorithm.

@@

Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue push operations.
As both appear in pairs, reducing the number of push operations, is beneficial for both.
Avoiding push operations on low degree nodes is the focus of this section.

@@

Our algorithm differs from A* when removing a node $x$ from the queue.
A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
If A* succeeds, $y$'s weight in the queue is set to $D[y]+h(y)$.
Our algorithm, however, behaves differently, if $d(y)\le 2$.
Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
Instead, it iteratively tries to reduce all $D[y_i]$.
If it does not reach $z$, then only $D$ is modified but no queue action is performed.
If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h(z)$.

@@

Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h(\beta)$.
Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h(b)$.
If $b$ respectively $\beta$ are not reached, our algorithm does nothing.

@@

A lot of nodes in road networks lead to dead-ends.
Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.

@@

A* is a flexible algorithm.
Many extended problems can therefore be solved using the CH-Potentials algorithm.
In this section, we describe some extended problems and how they can be solved with CH-Potentials.

@@

Avoiding tunnels and/or highways is a common feature of navigation devices.
Implementing this feature with CH-Potentials is easy.
We set $w_\ell$ to the freeflow travel time.
If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
Otherwise, $w_q$ is set to the freeflow travel time.

@@

Truck routing differs from car routing due to night driving bans and other restrictions.
In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is used is such scenario.
The work considers time-dependent blocked edges and waiting at parking locations.
Further, a trade-off between arrival time and route quality is considered.
